\vspace{-0.2in} Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-06 main.tex
Tweaked figures, applied changes from Hossein.

\section{Introduction} 
\vspace{-0.05in}
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

\subsection{Literature Review} \vspace{-0.05in}
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}
\vspace{-0.05in}
In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

\vspace{-0.05in}
\section{Hardware Concurrency} \label{sec:concurrency}
\vspace{-0.05in}
In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

@@

\subsection{CPU Hardware Concurrency}
\vspace{-0.05in}
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds. The baseline architecture for our experiments 
is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM exceeds the amount used
by the algorithms, and is not an important performance consideration.

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.05in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SM), each of which is roughly analogous to a CPU core.
For Fermi architecture GPUs there are up to 16 SMs, and each SM is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

\section{Parallelism in the Face Alignment Stage}\vspace{-0.05in}
\label{sec:alignment}

@@

\subsection{CPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_cpu}

@@

\subsection{GPU Implementation} \vspace{-0.05in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:gpu_caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

@@

\section{Parallelization of the Face Recognition Stage} \vspace{-0.05in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients in $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} \vspace{-0.05in}
Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} \vspace{-0.05in}
\label{sec:experiment} 
In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.
%To generate a meaningful comparison between different hardware architectures,
%in all cases we compare the performance of hardware architectures on a
%per-board basis:  For CPU implementations, the benchmark makes use of all of
%the cores in as many CPU's are present.  For GPU implementations, the benchmark
%makes the best use of the entire GPU chip (most GPU boards have a single GPU
%chip).

@@

\subsection{Face Recognition Pipeline Benchmark} \vspace{-0.05in}
\label{sec:benchmark}
This section presents benchmarks of the CPU and GPU implementations of the
alignment stage (i.e., Algorithm
\ref{alg:iterative_alignment}) as well as the recognition stage (i.e., Algorithm \ref{alg:alm_rec}) of the
face recognition pipeline.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=3.3in]{figures/alignment_runtime_graph}\vspace{0.05in}
\caption{Alignment stage runtime vs. size of training database.} \vspace{-0.1in}
\label{fig:alignment_stage_runtime}
\end{figure}

@@

Figure \ref{fig:recognition_stage_runtime} shows the recognition stage runtime
for canonical images of size: $32\times32$, $48 \times 48$, $64 \times 64$, $96 \times
96$, and $128 \times 128$.  For all image resolutions, the problem size
is sufficiently small that the CPU is significantly faster than the GPU.  Note
also that for both implementations, the recognition stage takes much less time
than the alignment stage.
\begin{figure}[t!]
\centering
\includegraphics[width=2.5in]{figures/speedVsResolution} \vspace{-0.05in}
\caption{Recognition stage runtime vs. face window resolution}\vspace{-0.1in}
\label{fig:recognition_stage_runtime} \end{figure}

Finally, we perform an experiment verifying the recognition accuracy of the
overall pipeline.  As shown in Figure \ref{fig:accuracy_vs_resolution},
at the optimal resolution, which happens to match the alignment stage
resolution, the GPU implementation reaches 95\% recognition rate, the max
achievable given the alignment selection process.  For significantly lower resolutions,
the accuracy drops off significantly.  The slight difference in CPU vs. GPU
accuracy may be a result of numerical precision differences in our matrix
inversion and vector reduction routines.
\begin{figure}[t!]
\centering
\includegraphics[width=2.5in]{figures/accuracyVsResolution} \vspace{-0.05in}
\caption{Recognition rate of the full pipeline vs. face window resolution.} \vspace{-0.1in}
\label{fig:accuracy_vs_resolution}
\end{figure}

\section{Conclusion}\vspace{-0.05in}
We have demonstrated that on both CPU and GPU algorithms, parallelizations of
ALM that solve multiple face alignment problems concurrently are significantly
faster than implementations that rely purely on vendor-supplied BLAS libraries.
Furthermore, thanks to a combination of faster hardware and more efficient use
of the hardware, we have demonstrated dramatic improvements in recognition
speed over previously reported implementations.
As CPU manufacturers increase the number of cores and vector widths, and as GPU 
manufacturers increase the amount
of cache, both architectures are rapidly converging towards an architectural balance
that increasingly favorable for $\ell_1$-min based face recognition.
% of the same pipeline.  Whereas
%earlier implementations required over a second to align a test image against
%just one of the 250 subjects in Multi-PIE session 1, the proposed 
%implementations can perform the same task in 
%{\bf under 10 seconds}!.  
%The parallelization techniques presented here make it feasible to implement an
%access control system for hundreds of subjects on a handful of purpose-build
%multi-GPU workstations.

@@

{\small
\bibliographystyle{ieee}
\bibliography{faces-short}
}

>>>>

\section{Introduction} 
\vspace{-0.06in}
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
The problem refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

\subsection{Literature Review} \vspace{-0.06in}
Traditionally, $\ell_1$-min (a.k.a.
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

@@

\section{Augmented Lagrangian Method}
\label{sec:ALM}
\vspace{-0.06in}
In this section, we briefly describe the ALM algorithm for $\ell_1$-min
\eqref{eq:l1min} \cite{YangA2010-ICIP} and analyze its complexity. Lagrange
multipliers have been frequently used in convex programming to eliminate
equality constraints via adding a penalty term to the cost function for
infeasible points. ALM methods differ from other penalty-based approaches by
simultaneously estimating the optimal solution and Lagrange multipliers in an
iterative fashion.  For $\ell_1$-min \eqref{eq:l1min}, the augmented Lagrange
function is defined as: 
\begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx) + \frac{\mu}{2} \| \bb-A\xx \|_2^2, 
\end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of Lagrange multipliers.

@@

\vspace{-0.06in}
\section{Hardware Concurrency} \label{sec:concurrency}
\vspace{-0.06in}
In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since caches
(regions of on-chip memory) are often orders of magnitude faster than off-chip
memory, their sizes and speeds have a dramatic effect on performance.  We give a
brief overview of the caches that are available in our target architectures,
and defer discussion of their performance effects to Sections
\ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

@@

\subsection{CPU Hardware Concurrency}
\vspace{-0.06in}
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
cache, and relatively high clock speeds. The baseline architecture for our experiments 
is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM exceeds the amount used
by the algorithms, and is not an important performance consideration.

@@

\subsection{GPU Hardware Concurrency} \vspace{-0.06in}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations, we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. 
%\footnote{A discussion of the CUDA programming model can be found in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SM), each of which is roughly analogous to a CPU core.
For Fermi architecture GPUs there are up to 16 SMs, and each SM is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

\section{Parallelism in the Face Alignment Stage}\vspace{-0.06in}
\label{sec:alignment}

@@

\subsection{CPU Implementation} \vspace{-0.06in}
\label{sec:alignment_implementation_cpu}

@@

\subsection{GPU Implementation} \vspace{-0.06in}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, cache is relatively unimportant on the GPU for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:gpu_caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is its ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

@@

\section{Parallelization of the Face Recognition Stage} \vspace{-0.06in}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subject classes with lowest
alignment error are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients in $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

\subsection{Recognition Stage Implementation} \vspace{-0.06in}
Compared to problem-level
parallelism, the exclusively pixel-level parallelism in Algorithm \ref{alg:alm_rec} is
relatively straightforward to exploit:  
On the CPU, most of the operations map
well onto MKL BLAS calls, and operations that do not can be easily
parallelized using OpenMP and auto-vectorization.

@@

\section{Experiments} \vspace{-0.06in}
\label{sec:experiment} 
In this section, we benchmark the
performance of our parallel implementations of $\ell_1$-min on CPU and GPU
platforms.  In order to show how our $\ell_1$-min algorithms scale with problem
size, in Section \ref{sec:simulation} we will begin with benchmarks for the
general $\ell_1$-min problem on synthetic data.  We will then progress in
Section \ref{sec:benchmark} to demonstrating the speed and accuracy of our
implementations as applied to the alignment and recognition stages of the face
recognition problem.
%To generate a meaningful comparison between different hardware architectures,
%in all cases we compare the performance of hardware architectures on a
%per-board basis:  For CPU implementations, the benchmark makes use of all of
%the cores in as many CPU's are present.  For GPU implementations, the benchmark
%makes the best use of the entire GPU chip (most GPU boards have a single GPU
%chip).

@@

\subsection{Face Recognition Pipeline Benchmark} \vspace{-0.06in}
\label{sec:benchmark}
This section presents benchmarks of the CPU and GPU implementations of the
alignment stage (i.e., Algorithm
\ref{alg:iterative_alignment}) as well as the recognition stage (i.e., Algorithm \ref{alg:alm_rec}) of the
face recognition pipeline.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=3.3in]{figures/alignment_runtime_graph}\vspace{0.06in}
\caption{Alignment stage runtime vs. size of training database.} \vspace{-0.1in}
\label{fig:alignment_stage_runtime}
\end{figure}

@@

Figure \ref{fig:recognition_stage_runtime} shows the recognition stage runtime
for canonical images of size: $32\times32$, $48 \times 48$, $64 \times 64$, $96 \times
96$, and $128 \times 128$.  For all image resolutions, the problem size
is sufficiently small that the CPU is significantly faster than the GPU.  Note
also that for both implementations, the recognition stage takes much less time
than the alignment stage.
\begin{figure}[t!]
\centering
\includegraphics[width=2.5in]{figures/speedVsResolution} \vspace{-0.06in}
\caption{Recognition stage runtime vs. face window resolution}\vspace{-0.1in}
\label{fig:recognition_stage_runtime} \end{figure}

Finally, we perform an experiment verifying the recognition accuracy of the
overall pipeline.  As shown in Figure \ref{fig:accuracy_vs_resolution},
at the optimal resolution, which happens to match the alignment stage
resolution, the GPU implementation reaches 95\% recognition rate, the max
achievable given the alignment selection process.  For significantly lower resolutions,
the accuracy drops off significantly.  The slight difference in CPU vs. GPU
accuracy may be a result of numerical precision differences in our matrix
inversion and vector reduction routines.
\begin{figure}[t!]
\centering
\includegraphics[width=2.5in]{figures/accuracyVsResolution} \vspace{-0.06in}
\caption{Recognition rate of the full pipeline vs. face window resolution.} \vspace{-0.1in}
\label{fig:accuracy_vs_resolution}
\end{figure}

\section{Conclusion}\vspace{-0.06in}
We have demonstrated that on both CPU and GPU algorithms, parallelizations of
ALM that solve multiple face alignment problems concurrently are significantly
faster than implementations that rely purely on vendor-supplied BLAS libraries.
Furthermore, thanks to a combination of faster hardware and more efficient use
of the hardware, we have demonstrated dramatic improvements in recognition
speed over previously reported implementations.
As CPU manufacturers increase the number of cores and vector widths, and as GPU 
manufacturers increase the amount
of cache, both architectures are rapidly converging towards an architectural balance
that increasingly favorable for $\ell_1$-min based face recognition.
% of the same pipeline.  Whereas
%earlier implementations required over a second to align a test image against
%just one of the 250 subjects in Multi-PIE session 1, the proposed 
%implementations can perform the same task in 
%{\bf under 10 seconds}!.  
%The parallelization techniques presented here make it feasible to implement an
%access control system for hundreds of subjects on a handful of purpose-build
%multi-GPU workstations.

@@

\vspace{-0.06in}
{\small
\bibliographystyle{ieee}
\bibliography{faces-short}
}
